---
title: "STA288 Final Project"
author: "Belal Hajjaj, Aaron Jin, Justin Yu, Brandon Luong, Modhar Al Qasser"
date: "2024-03-27"
geometry: margin = 0.5in
output: 
  pdf_document: default
---


## Author Contributions
BH, AJ, JY, BL, and MA - contributed to the development of the methodology  
BH, AJ, JY, BL, and MA - contributed to the data collection  
JY - contributed to R-markdown files
JY - contributed to statistical analysis of the data  
JY - contributed to drafting and writing the report  

## Introduction

## Materials and Methods
Participants were sampled from the village of Arcadia on the island of Providence. Arcadia was chosen for its large population size of 4339, which encompasses a larger range of individuals than other villages, improving generalizability. Since there was no feasible method of obtaining a list of all individuals to take a simple random sample, a multistage sampling strategy was employed. Only adults at least 19 years of age were included to ensure participants were old enough to drink and complete study tasks.

Excel functions were used to generate a list of 100 house numbers out of the 1571 houses in Arcadia. From each selected house, the adults were numbered in order of appearance. R `runif()` was then used to generate a random number, and the corresponding resident was selected.

Participants who did not provide consent, empty houses, and duplicates were removed from the data, resulting in a final sample size of n = 90.

## ANALYSIS (TEMPORARY)

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```
**Import data**
```{r}
raw <- read_csv("data.csv")
```
**Standardize data formatting**
```{r}
data <- raw %>%
  # keep only entries with data
  filter(Alcohol_consumption != "DUPLICATE" & Alcohol_consumption != "EMPTY" & Alcohol_consumption != "NO CONSENT") %>% 
  # clean up formatting
  mutate(Name = str_to_title(Name), # change all names to title case
        
        # standardize alcohol consumption format
        Alcohol_consumption = case_when(
          str_detect(Alcohol_consumption, fixed("several times each season", ignore_case = T)) ~ "Several times each season",
          str_detect(Alcohol_consumption, fixed("once or twice each season", ignore_case = T)) ~ "Once or twice each season",
          str_detect(Alcohol_consumption, fixed("once or twice a year", ignore_case = T)) ~ "Once or twice a year",
          str_detect(Alcohol_consumption, fixed("couple of times each day", ignore_case = T)) ~ "Couple of times each day",
          str_detect(Alcohol_consumption, fixed("drink each day", ignore_case = T)) ~ "Once each day"
        ),
        
        # standardize forgetfulness format
        Forgetful = case_when(
          str_detect(Forgetful, fixed("not at all", ignore_case = T)) ~ "Not at all",
          str_detect(Forgetful, fixed("a little", ignore_case = T)) ~ "A little",
          str_detect(Forgetful, fixed("moderately", ignore_case = T)) ~ "Moderately",
          str_detect(Forgetful, fixed("moderatly", ignore_case = T)) ~ "Moderately" # typo in one entry
        )
  )
```

**Preliminary Analyses**
*TEMPORARY: These results will be summarized using side-by-side boxplots, numerical summaries (mean,
median, standard deviation, range), histograms, and stacked bar charts. The numerical summaries will
provide a general overview of the center and spread/variation among the memory tests. The side-by-side
boxplots will allow us to visualize the central tendencies and the spread of the memory tests among each
group. Furthermore, histograms will allow us to analyze the shape of the results of the memory tests.
The qualitative response variable will be visualized and analyzed through the stacked bar chart as it allows
us to make comparisons across each group.*

```{r}
# side by side histograms

# cards
data %>% 
  # order by increasing consumption
  mutate(Alcohol_consumption = factor(Alcohol_consumption, levels=c("Once or twice a year", "Once or twice each season", "Several times each season", "Once each day", "Couple of times each day"))) %>% 
  ggplot(aes(x = Alcohol_consumption, y = Cards)) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 10, vjust = 0.5))

# vocab
data %>% 
  # order by increasing consumption
  mutate(Alcohol_consumption = factor(Alcohol_consumption, levels=c("Once or twice a year", "Once or twice each season", "Several times each season", "Once each day", "Couple of times each day"))) %>% 
  ggplot(aes(x = Alcohol_consumption, y = Vocab)) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 10, vjust = 0.5))

# mosaic of forgetfulness (totally unreadable)
# create two-way table
mostable <- table(data$Alcohol_consumption, data$Forgetful)
mostable

# create mosaic plot
mosaicplot(mostable, main = "Self-reported Forgetfulness by Alcohol Consumption Frequency", las=2)

# replace with ggplot
```
**Hypothesis tests**
```{r}
anova_cards <- aov(data$Cards ~ data$Alcohol_consumption)
summary(anova_cards)

anova_vocab <- aov(data$Vocab ~ data$Alcohol_consumption)
summary(anova_vocab)

chisq.test(data$Alcohol_consumption, data$Forgetful)
```

## Results

## Conclusions

## Discussion

## References

## Appendix